---
title: "Modeling effect of peaker plant operations on air quality in Pittsfield, MA"
author: "Anna He"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
The purpose of this script is to analyze the effect of peaker power plant operations on air quality (PM2.5, PM10, and NO2 concentrations) at two locations: 
Allendale Elementary School in Pittsfield, MA and the population-weighted centroid in Pittsfield, MA. 

Air quality data from BEAT and peaker plant operations data from the EPA CAMPD were downloded for April 1, 2024 - September 30, 2025.
The time period of analysis was Summer 2024 and Summer 2025 (June, July, and August).

```{r load-libraries, warning = F, echo = F}
library(ggplot2)
library(pheatmap)
library(cowplot)
library(sf)
library(dplyr)
library(broom) 
library(spdep)
library(splines) 
library(CARBayes) # Fitting conditional autoregressive (CAR) models
library(tidyr) 
library(ggmap)
library(data.table)
library(tidyverse)
library(spatstat)
library(tinytex)
library(lubridate)
library(gstat)
library(stars)
library(terra)
library(cowplot)

```

## Loading the input data

```{r load-data, echo=T}
#time series air quality data from BEAT
sensorsaq_all <- fread("data/input/sensordata_clean.csv") 

#weather data (includes hourly temp, wind direction, and windspeed) from National Centers for Environmental Information 
weather_all <- fread("data/input/WeatherClean.csv") 

#sensor location and description data from BEAT
sensorinfo <- fread("data/input/sensors.csv")

#peaker plant operations data from the EPA (cleaned)
peakerops <- fread("data/input/peakerop_clean.csv") 

``` 

#PART 1: JOINING AND VISUALIZING AIR QUALITY DATA 

```{r 1-join-sensor-info, warning = FALSE, echo = TRUE}
#join sensor location/name info to the time series sensor air quality data
sensors_joined_1 <- dplyr::left_join(sensorsaq_all, sensorinfo, 
                                     by = c("datasourceId" = "datasourceID"))

head(sensors_joined_1)

```
```{r 2-join-peaker-operations, warning = FALSE, echo = TRUE}
#join the peaker operations info to sensors air quality data
sensors_joined_2 <- dplyr::left_join(sensors_joined_1, peakerops, 
                                     by = c("StartOfPeriod" = "datetime"))

head(sensors_joined_2)

```

```{r 3a-check-weather-data, warning = F, echo = T}
#there are multiple report types, SOM, SOD (summary of month and summary of day) 
#with null values. what happens if I remove them?

weather_filtered <- weather_all |> dplyr::filter(REPORT_TYPE == "FM-15")
head(weather_filtered)
class(weather_filtered$HourlyWindDirection)
table(weather_filtered$HourlyWindDirection)

#There are 1041 values with variable wind direction. Should I force to be NAs?
#decided not to include wind direction as a variable in the model

```

```{r 3b-join-weather-data, warning = FALSE, echo = TRUE}
#Join the weather data to sensors air quality data
#Got an error that there were multiple weather dates for 2025;03;08;21
weather_test <- weather_filtered |> dplyr::filter(DATE == "2025;03;08;21")

#dropped the second row because both measurements were taken at the same time based on the original weather data
weather_filtered_unique <- weather_filtered[-8201,]

sensors_joined_3 <- dplyr::left_join(sensors_joined_2, weather_filtered_unique, by = c("StartOfPeriod" = "DATE"))

#sensors_joined_3 is the full dataset of sensor air quality data joined to sensor information
#peaker plant operation information
#and weather information

head(sensors_joined_3)

write.csv(sensors_joined_3, "data/output/sensors_joined_all.csv")
```
```{r 4-filter-to-summer, warning = FALSE, echo = TRUE}
#filter the complete dataset to just summer 2024 and summer 2025
sensors_summer_1 <- sensors_joined_3 |> dplyr::filter(season == "Summer")

head(sensors_summer_1)

``` 

```{r 5-fireworks, warning = FALSE, echo = TRUE}
#add a new column, fireworks, to represent potential July 4th covariate 
#dates that fall on or after July 4th on a weekend may have elevated pollutants due to fireworks
#if the date falls on the following dates, assign it a value of 1 for fireworks; otherwise, 0

sensors_summer_2 <- sensors_summer_1 

sensors_summer_2$fireworks <- ifelse(sensors_summer_2$date %in% c("7/4/2024", "7/5/2024", "7/6/2024", "7/7/2024", "7/4/2025", "7/5/2025", "7/6/2025"), "1", "0")
```

```{r 6-wildfire-smoke, warning = FALSE, echo = TRUE}
#add a new column, smoke, to represent potential covariate from wildfire smoke from Canada
#check for dates that air quality alerts were issued in summer 2024 and summer 2025 due to wildfire smoke
#sources: local news articles, Mass DEP, Berkshire health orgs
#if the date falls on the following dates with an air quality alert due to smoke, assign it a value of 1; otherwise 0

sensors_summer_2$smoke <- ifelse(sensors_summer_2$date %in% c("8/16/2024", "6/5/2025", "7/26/2025", "7/27/2025", "8/4/2025", "8/5/2025", "8/6/2025"), "1", "0")

head(sensors_summer_2)
```

```{r 7-filter-to-allendale-elementary-school, warning = FALSE, echo = TRUE}
allendale <- sensors_summer_2 |> dplyr::filter(name == "Allendale Elementary School")

#check for NAs and distribution
table(allendale$no2Conc1HourMean.status)
table(allendale$pm2_5ConcMass1HourMean.status)
table(allendale$pm10ConcMass1HourMean.qcAssessment)

sum(is.na(allendale$no2Conc1HourMean.value))
sum(is.na(allendale$pm2_5ConcMass1HourMean.value))
sum(is.na(allendale$pm10ConcMass1HourMean.raw))
sum(is.na(allendale$HourlyDryBulbTemperature))
sum(is.na(allendale$HourlyRelativeHumidity))
sum(is.na(allendale$HourlyWindDirection))
sum(is.na(allendale$HourlyWindSpeed))

#all important variables have less than 5% NAs.
#only 0.06% (3/4416) of observations are missing for pollutant concentrations
#2.79% (123/4416) of observations are missing for temperature data
#3.33% (147/4416) of observations are missing for humidity data
#1.36% (60/4416) of observations are missing for hourly wind direction and wind speed

#check for NAs distribution
allendale$date[is.na(allendale$no2Conc1HourMean.value)]
allendale$date[is.na(allendale$pm10ConcMass1HourMean.raw)]
allendale$date[is.na(allendale$pm2_5ConcMass1HourMean.value)]

allendale$date[is.na(allendale$HourlyDryBulbTemperature)]
allendale$hour[is.na(allendale$HourlyDryBulbTemperature)]

hist(allendale$hour[is.na(allendale$HourlyDryBulbTemperature)])



```

```{r 8a-initial-visualization-pm2.5}
#boxplot comparing PM2.5 levels when the power plant is operating or not operating
pm25box <- ggplot(data = allendale, aes(x = factor(operating), y = pm2_5ConcMass1HourMean.value)) + 
  geom_boxplot(na.rm = TRUE) +
  labs(title = "Mean hourly PM2.5 levels (µg/m3)",
       x = "Operating (0 = No, 1 = Yes)",
       y = "Concentration (µg/m3)") +
  theme_minimal()+ 
  theme(plot.title = element_text(size = 40),
        axis.title.x = element_text(size = 30), 
        axis.title.y = element_text(size = 30),
        axis.text = element_text(size = 25))


#factor by smoke
ggplot(data = allendale, aes(x = factor(smoke), y = pm2_5ConcMass1HourMean.value)) + 
  geom_boxplot(na.rm = TRUE) +
  labs(title = "Mean hourly PM2.5 levels at Allendale Elementary School \nSummer 2024 and 2025",
       x = "Smoke alerts (0 = No, 1 = Yes)",
       y = "PM2.5 Concentration  (micrograms/m³)") +
  theme_minimal()

#factor by fireworks
ggplot(data = allendale, aes(x = factor(fireworks), y = pm2_5ConcMass1HourMean.value)) + 
  geom_boxplot(na.rm = TRUE) +
  labs(title = "Mean hourly PM2.5 levels at Allendale Elementary School \nSummer 2024 and 2025",
       x = "Smoke alerts (0 = No, 1 = Yes)",
       y = "PM2.5 Concentration  (micrograms/m³)") +
  theme_minimal()

#factor by hour
ggplot(data = allendale, aes(x = factor(hour), y = pm2_5ConcMass1HourMean.value)) + 
  geom_boxplot(na.rm = TRUE) +
  labs(title = "Mean hourly PM2.5 levels at Allendale Elementary School \nSummer 2024 and 2025",
       x = "Hour",
       y = "PM2.5 Concentration  (micrograms/m³)") +
  theme_minimal()

#factor by weekday
ggplot(data = allendale, aes(x = factor(weekday_num), y = pm2_5ConcMass1HourMean.value)) + 
  geom_boxplot(na.rm = TRUE) +
  labs(title = "Mean hourly PM2.5 levels at Allendale Elementary School \nSummer 2024 and 2025",
       x = "Weekday (0=No, 1 = Yes)",
       y = "PM2.5 Concentration  (µg/m3)") +
  theme_minimal()


#scatterplot of PM2.5 over time
class(allendale$date)
allendale$date <- as.Date(allendale$date, format = "%m/%d/%Y")
class(allendale$date)
allendale$year <- format(allendale$date, "%Y")

allendale$operating_f <- factor(allendale$operating,
                                levels = c(0, 1),
                                labels = c("Not operating", "Operating"))


#plot scatterplot of data faceted by year and colored by operational status
pm25date<- ggplot(data = allendale, aes(x = date, 
                             y = pm2_5ConcMass1HourMean.value, 
                             color = operating_f)) +
  geom_point(alpha = 0.4, size = 1) +
  facet_wrap(~year, scales = "free_x") +
  labs(title = "Mean hourly PM2.5 levels at Allendale Elementary", 
       x = "Date",
       y = "Concentration (µg/m3)",
       color = "Power plant status") +
  theme_minimal(base_size = 16)


ggsave("plots/pm25date.png", pm25date, width = 8, height = 6, units = "in")

#plot scatterplot of data faceted by hour and colored by operational status
#don't use, not very helpful
ggplot(data = allendale, aes(x = hour, 
                             y = pm2_5ConcMass1HourMean.value, 
                             color = operating_f)) +
  geom_point(alpha = 0.4) +
  facet_wrap(~year, scales = "free_x") +
  labs(title = "Hourly PM2.5 Levels at Allendale Elementary School", 
       x = "Hour",
       y = "PM2.5 (micrograms/cubic meters)",
       color = "Power plant status") +
  theme_minimal()



#plot scatterplot over datetime
allendale$datetime <- as.POSIXct(
  paste(allendale$date, allendale$hour),
  format = "%Y-%m-%d %H")

ggplot(data = allendale, aes(x = datetime, 
                             y = pm2_5ConcMass1HourMean.value, 
                             color = operating_f)) +
  geom_point(alpha = 0.4) +
  geom_line(alpha = 0.4) +
  facet_wrap(~year, scales = "free_x") +
  labs(title = "Hourly PM2.5 Levels at Allendale Elementary School", 
       x = "Date",
       y = "PM2.5 (micrograms/cubic meters)",
       color = "Power plant status") +
  theme_minimal()

summary(allendale$pm2_5ConcMass1HourMean.value)
#allendale |> dplyr::group_by(operating)

``` 

```{r 8b-visualization-no2}
#boxplot comparing NOX levels when the power plant is operating or not operating
no2box <- ggplot(data = allendale, aes(x = factor(operating), y = no2Conc1HourMean.value)) + 
  geom_boxplot(na.rm = TRUE) +
  labs(title = "Mean hourly NO2 levels (ppb)",
       x = "Operating (0 = No, 1 = Yes)",
       y = "Concentration (ppb)") +
  theme_minimal()+ 
  theme(plot.title = element_text(size = 40),
        axis.title.x = element_text(size = 30), 
        axis.title.y = element_text(size = 30),
        axis.text = element_text(size = 25))



no2box

no2date<- ggplot(data = allendale, aes(x = date, 
                             y = no2Conc1HourMean.value, 
                             color = operating_f)) +
  geom_point(alpha = 0.4, size = 1) +
  facet_wrap(~year, scales = "free_x") +
  labs(title = "Mean hourly NO2 levels at Allendale Elementary", 
       x = "Date",
       y = "Concentration (ppb)",
       color = "Power plant status") +
  theme_minimal(base_size = 16)


ggsave("plots/no2date.png", no2date, width = 8, height = 6, units = "in")

```

```{r 8c-visualization-pm10}
#boxplot comparing NOX levels when the power plant is operating or not operating
pm10box <- ggplot(data = allendale, aes(x = factor(operating), y = allendale$pm10ConcMass1HourMean.raw)) + 
  geom_boxplot(na.rm = TRUE) +
  labs(title = "Mean hourly PM10 levels (µg/m3)",
       x = "Operating (0 = No, 1 = Yes)",
       y = "Concentration (µg/m3)") +
  theme_minimal() + 
  theme(plot.title = element_text(size = 40),
        axis.title.x = element_text(size = 30), 
        axis.title.y = element_text(size = 30),
        axis.text = element_text(size = 25))


pm10box


operatingbox <- cowplot::plot_grid(pm25box, pm10box, no2box)
cowplot::save_plot("plots/operatingboxplot.png", operatingbox, base_height = 21, ncol = 1)

pm10date<- ggplot(data = allendale, aes(x = date, 
                             y = allendale$pm10ConcMass1HourMean.raw, 
                             color = operating_f)) +
  geom_point(alpha = 0.4, size = 1) +
  facet_wrap(~year, scales = "free_x") +
  labs(title = "Mean hourly PM10 levels at Allendale Elementary", 
       x = "Date",
       y = "Concentration (µg/m3)",
       color = "Power plant status") +
  theme_minimal(base_size = 16)


ggsave("plots/pm10date.png", pm10date, width = 8, height = 6, units = "in")

```

```{r 8d-statistics-allendale}
#pm2.5 statistics at allendale
mean(allendale$pm2_5ConcMass1HourMean.value, na.rm = TRUE)
median(allendale$pm2_5ConcMass1HourMean.value, na.rm = TRUE)
quantile(allendale$pm2_5ConcMass1HourMean.value, na.rm = TRUE)
sd(allendale$pm2_5ConcMass1HourMean.value, na.rm = TRUE)

#pm10 statistics at allendale
mean(allendale$pm10ConcMass1HourMean.raw, na.rm = TRUE)
median(allendale$pm10ConcMass1HourMean.raw, na.rm = TRUE)
quantile(allendale$pm10ConcMass1HourMean.raw, na.rm = TRUE)
sd(allendale$pm10ConcMass1HourMean.raw, na.rm = TRUE)

#no2 statistics at allendale
mean(allendale$no2Conc1HourMean.value, na.rm = TRUE)
median(allendale$no2Conc1HourMean.value, na.rm = TRUE)
quantile(allendale$no2Conc1HourMean.value, na.rm = TRUE)
sd(allendale$no2Conc1HourMean.value, na.rm = TRUE)

#filter peaker power plant data when power plant is on vs. off
allendaleon <- allendale |>filter(operating == 1)
allendaleoff <- allendale |> filter(operating == 0)

#statistics for pm2.5 at allendale when peaker is on
mean(allendaleon$pm2_5ConcMass1HourMean.value, na.rm = TRUE)
median(allendaleon$pm2_5ConcMass1HourMean.value, na.rm = TRUE)
quantile(allendaleon$pm2_5ConcMass1HourMean.value, na.rm = TRUE)
sd(allendaleon$pm2_5ConcMass1HourMean.value, na.rm = TRUE)

#statistics for pm10 at allendale when peaker is on
mean(allendaleon$pm10ConcMass1HourMean.raw, na.rm = TRUE)
median(allendaleon$pm10ConcMass1HourMean.raw, na.rm = TRUE)
quantile(allendaleon$pm10ConcMass1HourMean.raw, na.rm = TRUE)
sd(allendaleon$pm10ConcMass1HourMean.raw, na.rm = TRUE)

#statistics for no2 at allendale when peaker is on
mean(allendaleon$no2Conc1HourMean.value, na.rm = TRUE)
median(allendaleon$no2Conc1HourMean.value, na.rm = TRUE)
quantile(allendaleon$no2Conc1HourMean.value, na.rm = TRUE)
sd(allendaleon$no2Conc1HourMean.value, na.rm = TRUE)

#statistics for pm2.5 at allendale when peaker is off
mean(allendaleoff$pm2_5ConcMass1HourMean.value, na.rm = TRUE)
median(allendaleoff$pm2_5ConcMass1HourMean.value, na.rm = TRUE)
quantile(allendaleoff$pm2_5ConcMass1HourMean.value, na.rm = TRUE)
sd(allendaleoff$pm2_5ConcMass1HourMean.value, na.rm = TRUE)

#statistics for pm10 at allendale when peaker is off
mean(allendaleoff$pm10ConcMass1HourMean.raw, na.rm = TRUE)
median(allendaleoff$pm10ConcMass1HourMean.raw, na.rm = TRUE)
quantile(allendaleoff$pm10ConcMass1HourMean.raw, na.rm = TRUE)
sd(allendaleoff$pm10ConcMass1HourMean.raw, na.rm = TRUE)

#statistics for no2 at allendale when peaker is off
mean(allendaleoff$no2Conc1HourMean.value, na.rm = TRUE)
median(allendaleoff$no2Conc1HourMean.value, na.rm = TRUE)
quantile(allendaleoff$no2Conc1HourMean.value, na.rm = TRUE)
sd(allendaleoff$no2Conc1HourMean.value, na.rm = TRUE)
```
```{r 8e-output-allendale.csv}
write.csv(allendale, "data/output/allendale_aq.csv")
```


#PART 2: RUNNING SIMPLE LINEAR MODELS FOR PM2.5, NO2, AND PM10

```{r 9a-run-lm-pm2.5}
#check variable classes
allendale$hour <- as.factor(allendale$hour)

#run first linear model
pm25lm1 <- lm(formula = pm2_5ConcMass1HourMean.value ~ operating_f + 
                hour + weekday_num, data = allendale)

summary(pm25lm1)

#add in more variables

pm25lm2 <- lm(formula = pm2_5ConcMass1HourMean.value ~ operating_f + 
                hour + weekday_num + fireworks + smoke, data = allendale)

summary(pm25lm2)

#add in more variables

pm25lm3 <- lm(formula = pm2_5ConcMass1HourMean.value ~ operating_f + 
                hour + year + weekday_num + fireworks + smoke, data = allendale)

summary(pm25lm3)

#add in more variables
allendale$HourlyWindSpeed <- as.integer(allendale$HourlyWindSpeed)

pm25lm4 <- lm(formula = pm2_5ConcMass1HourMean.value ~ operating_f + 
                hour + year + weekday_num + fireworks + smoke + HourlyDryBulbTemperature, 
              data = allendale)

summary(pm25lm4)

#lm5
pm25lm5 <- lm(formula = pm2_5ConcMass1HourMean.value ~ operating_f + 
                hour + year + weekday_num + fireworks + smoke + HourlyDryBulbTemperature + 
                HourlyRelativeHumidity, data = allendale)

summary(pm25lm5)

#lm6

pm25lm6 <- lm(formula = pm2_5ConcMass1HourMean.value ~ operating_f + 
                hour + year + weekday_num + fireworks + smoke + HourlyDryBulbTemperature + 
                HourlyRelativeHumidity + HourlyWindSpeed, data = allendale)

summary(pm25lm6)

```

```{r 9b-run-linear-models-no}

#run first linear model
no2lm1 <- lm(formula = no2Conc1HourMean.value ~ operating_f + 
               hour + weekday_num, data = allendale)

summary(no2lm1)

#add in more variables

no2lm2 <- lm(formula = no2Conc1HourMean.value ~ operating_f + 
               hour + weekday_num + fireworks + smoke, data = allendale)

summary(no2lm2)

#add in more variables

no2lm3 <- lm(formula = no2Conc1HourMean.value ~ operating_f + 
               hour + year + weekday_num + fireworks + smoke, data = allendale)

summary(no2lm3)

#add in more variables

no2lm4 <- lm(formula = no2Conc1HourMean.value ~ operating_f + 
               hour + year + weekday_num + fireworks + smoke + HourlyDryBulbTemperature, 
             data = allendale)

summary(no2lm4)

#lm5
no2lm5 <- lm(formula = no2Conc1HourMean.value ~ operating_f + 
               hour + year + weekday_num + fireworks + smoke + HourlyDryBulbTemperature + 
               HourlyRelativeHumidity, data = allendale)

summary(no2lm5)

#lm6

no2lm6 <- lm(formula = no2Conc1HourMean.value ~ operating_f + 
               hour + year + weekday_num + fireworks + smoke + HourlyDryBulbTemperature + 
               HourlyRelativeHumidity + HourlyWindSpeed, data = allendale)

summary(no2lm6)

```

```{r 9c-run linear models for pm10}
#run first linear model
pm10lm1 <- lm(formula = pm10ConcMass1HourMean.raw ~ operating_f + 
                hour + weekday_num, data = allendale)

summary(pm10lm1)

#add in more variables

pm10lm2 <- lm(formula = pm10ConcMass1HourMean.raw ~ operating_f + 
                hour + weekday_num + fireworks + smoke, data = allendale)

summary(pm10lm2)

#add in more variables

pm10lm3 <- lm(formula = pm10ConcMass1HourMean.raw ~ operating_f + 
                hour + weekday_num + fireworks + smoke + year, data = allendale)

summary(pm10lm3)

#add in more variables

pm10lm4 <- lm(formula = pm10ConcMass1HourMean.raw ~ operating_f +
                hour + year + weekday_num + fireworks + smoke + year + HourlyDryBulbTemperature, 
              data = allendale)

summary(pm10lm4)

#lm5
pm10lm5 <- lm(formula = pm10ConcMass1HourMean.raw ~ operating_f + 
                hour + year + weekday_num + fireworks + smoke + year + HourlyDryBulbTemperature + 
                HourlyRelativeHumidity, data = allendale)

summary(pm10lm5)

#lm6

pm10lm6 <- lm(formula = pm10ConcMass1HourMean.raw ~ operating_f + 
                hour + year + weekday_num + fireworks + smoke + year + HourlyDryBulbTemperature + 
                HourlyRelativeHumidity + HourlyWindSpeed, data = allendale)

summary(pm10lm6)
```

```{r 10a-select linear models }
#compare linear model results without weather data (lm3) and with weather data (lm6)
summary(pm25lm3)
summary(pm25lm6)

summary(pm10lm3)
summary(pm10lm6)

summary(no2lm3)
summary(no2lm6)
```

```{r 10b-plot-residuals-of-models}
#plot model residuals
plot(resid(pm25lm3))
plot(resid(pm10lm3))
plot(resid(no2lm3))

plot(pm25lm3)
plot(pm10lm3)
plot(no2lm3)


plot(pm25lm6)
plot(pm10lm6)
plot(no2lm6)


#they don't look the most normal...
```


```{r 10b-log transforming data running models}
#does log transforming the models change the model significance we see? 
#it does not

#pm2.5
pm25lm3log <- lm(formula = log(pm2_5ConcMass1HourMean.value) ~ operating_f + 
                   hour + year + weekday_num + fireworks + smoke, data = allendale)
summary(pm25lm3log)

pm25lm6log <- lm(formula = log(pm2_5ConcMass1HourMean.value) ~ operating_f + 
                   hour + year + weekday_num + fireworks + smoke + HourlyDryBulbTemperature + HourlyRelativeHumidity + HourlyWindSpeed, data = allendale)
summary(pm25lm6log)

#PM10
pm10lm3log <- lm(formula = log(pm10ConcMass1HourMean.raw) ~ operating_f + 
                   hour + year + weekday_num + fireworks + smoke, data = allendale)
summary(pm10lm3log)

pm10lm6log <- lm(formula = log(pm10ConcMass1HourMean.raw) ~ operating_f + 
                   hour + year + weekday_num + fireworks + smoke + HourlyDryBulbTemperature + 
                   HourlyRelativeHumidity + HourlyWindSpeed, data = allendale)
summary(pm10lm6log)

#no2
no2lm3log <- lm(formula = log(allendale$no2Conc1HourMean.value + 1) ~ operating_f + 
                  hour + year + weekday_num + fireworks + smoke, data = allendale)
summary(no2lm3log)

no2lm6log <- lm(formula = log(no2Conc1HourMean.value + 1) ~ operating_f + 
                  hour + year + weekday_num + fireworks + smoke + HourlyDryBulbTemperature + 
                  HourlyRelativeHumidity + HourlyWindSpeed, data = allendale)
summary(no2lm6log)


```

#PART 3: PREPARING DATA FOR INTERPOLATION

```{r 11- prepare-data-for-interpolation}
#convert classes for all sensors not just allendale
sensors_summer_2$hour <- as.factor(sensors_summer_2$hour)
sensors_summer_2$HourlyWindSpeed <- as.integer(sensors_summer_2$HourlyWindSpeed)
sensors_summer_2$date <- as.Date(sensors_summer_2$date, format = "%m/%d/%Y")
sensors_summer_2$year <- format(sensors_summer_2$date, "%Y")
sensors_summer_2$operating_f <- factor(sensors_summer_2$operating,
                                levels = c(0, 1),
                                labels = c("Not operating", "Operating"))
sensors_summer_2$datetime <- as.POSIXct(
  paste(sensors_summer_2$date, sensors_summer_2$hour),
  format = "%Y-%m-%d %H")

#write to a csv. updated December 13.
write.csv(sensors_summer_2, "data/output/summersensors.csv")

```


```{r 12-create-week-column}
#prepare data for interpolation
#create a sequential week column of every 7 days
sensors_summer_2$date <- as.Date(sensors_summer_2$date, format = "%m/%d/%Y")

start <- min(sensors_summer_2$date)

sensors_summer_2$week <- factor(
  floor(as.numeric(sensors_summer_2$date - start) / 7) + 1
)

table(sensors_summer_2$week)
#i see low numbers of observations in week 14 and week 66 because they fall at the tail end. 
#recode sensor weeks 14 to be part of 13 and week 66 to be part of 65
sensors_summer_2$week <- recode(sensors_summer_2$week,
                                "14" = "13",
                                "66" = "65")

table(sensors_summer_2$week)

#write to a csv.
write.csv(sensors_summer_2, "data/output/summersensors.csv")


``` 

```{r 13-weekly-avgno2-when-operating}
#calculate the average hourly no2 per week for each sensor when the power plant is on
#keep the longitude and latitude columns
#if any of those weeks contains a value of 1 for fireworks 
#(indicating on or around july 4)
#put a value of 1 for fireworks2

#if any of those weeks contains a value of 1 for smoke 
#(indicating air quality alert due to wildfire smoke
#put a value of 1 for smoke2)

no2_weekly_on <- sensors_summer_2 |> dplyr::filter(operating == 1,
                                                   !is.na(no2Conc1HourMean.value)) |>
  group_by(week, datasourceId) |>
  summarise(
    lon = first(longitude),
    lat = first(latitude),
    no2mean = mean(no2Conc1HourMean.value, na.rm = TRUE),
    fireworks2 = ifelse(any(fireworks == 1), 1, 0),
    smoke2 = ifelse(any(smoke == 1), 1, 0),
    temp = mean(HourlyDryBulbTemperature),
    humidity = mean(HourlyRelativeHumidity),
    windspeed = mean(HourlyWindSpeed),
    n_obs = n(),
    .groups = "drop"
  )

table(no2_weekly_on$week)

no2_weekly_on$operating = "Operating"

no2_weekly_on$year = ifelse(no2_weekly_on$week %in% c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13"), "2024", "2025")


```

```{r 14-no2-when-off}
#calculate the average hourly no2 per week for each sensor when the power plant is OFF
#keep the longitude and latitude columns
#if any of those weeks contains a value of 1 for fireworks 
#(indicating on or around july 4)
#put a value of 1 for fireworks2
#if any of those weeks contains a value of 1 for smoke 
#(indicating air quality alert due to wildfire smoke
#put a value of 1 for smoke2)
no2_weekly_off <- sensors_summer_2 |> dplyr::filter(operating == 0,
                                                   !is.na(no2Conc1HourMean.value)) |>
  group_by(week, datasourceId) |>
  summarise(
    lon = first(longitude),
    lat = first(latitude),
    no2mean = mean(no2Conc1HourMean.value, na.rm = TRUE),
    fireworks2 = ifelse(any(fireworks == 1), 1, 0),
    smoke2 = ifelse(any(smoke == 1), 1, 0),
    temp = mean(HourlyDryBulbTemperature),
    humidity = mean(HourlyRelativeHumidity),
    windspeed = mean(HourlyWindSpeed),
    n_obs = n(),
    .groups = "drop"
  )

table(no2_weekly_off$week)

no2_weekly_off$operating = "Not operating"
no2_weekly_off$year = ifelse(no2_weekly_off$week %in% c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13"), "2024", "2025")

#bind rows
no2 <- bind_rows(no2_weekly_on, no2_weekly_off)

write.csv(no2, "data/output/no2.csv")
```

#PART 4: INTERPOLATION
```{r 15-prepare data for interpolation}
#no2 is the data that we want to use for interpolation

#convert no2 to a spatial object
no2_sf <-  st_as_sf(no2, coords = c("lon", "lat"), crs=4326)
no2_26986 <- st_transform(no2_sf, crs = 26986)

#get pittsfield boundaries and other shapefiles
pittsfield26986 <- sf::read_sf("data/input/pittsfield26986/pittsfield26986.shp")
peaker26986 <- sf::read_sf("data/input/peaker26986/peaker26986.shp")
sensors26986 <- sf::read_sf("data/input/sensors26986/sensors26986.shp")
bg26986 <- sf::read_sf("data/input/pittsfieldbg26986/pittsfieldbg26986.shp")
airport26986 <- sf::read_sf("data/input/airport26986/airport26986.shp")
majroads26986 <- sf::read_sf("data/input/majroads26986/majroads26986.shp")

#get population-weighted centroid
centroid4326 <- st_sf(
  geometry = st_sfc(st_point(c(-73.24816089, 42.45187495))),
  crs = st_crs(4326)  # make sure CRS matches the raster
)
centroid26986 <- st_transform(centroid4326, st_crs(pittsfield26986))

#plot some points. COME BACK TO THIS
ggplot(pittsfield26986) + geom_sf(data = pittsfield26986) + 
geom_sf(data = peaker26986, color = "black", size = 3) +
  geom_sf(data = centroid26986, color = "blue", size = 5) +
  geom_sf(data = sensors26986, aes(color = name), shape = 21, size = 3) +
  labs(title = "Pittsfield City Boundary, sensors, and peaker plant")



#create a 500m grid 
grid500 <- sf::st_bbox(pittsfield26986) |> st_as_stars(dx=500) |>
  st_crop(pittsfield26986)

plot(grid500)

#create a 250m resolution grid
grid250 <- sf::st_bbox(pittsfield26986) |> st_as_stars(dx=250) |>
  st_crop(pittsfield26986)

plot(grid250)

#test 2 using the terra package
#pittsfield_boundary <- pittsfield26986 |> st_cast("MULTILINESTRING")
#pittsfield_boundary_vect <- terra::vect(pittsfield_boundary)

#pittsfield_raster500 <- terra::rast(
 # ext(pittsfield_boundary_vect),
  #resolution = 500,
  #crs = st_crs(pittsfield_boundary)$proj4string
#)

#hasValues(pittsfield_raster500)

#raster500 <- as.points(pittsfield_raster500, values = FALSE)


``` 

##test interpolating a raster and extracting data at a point with dummy data
```{r 16-interpolation-test}
#test idw for no2, no filtering
#this is with dummy data so won't be scientifically accurate
idw_test1 <- idw(formula = no2mean ~1,
                 locations = no2_26986 ,
                 newdata = grid250,
                 idp = 1)

names(idw_test1)

plot(idw_test1)["var1.pred"]

summary(idw_test1$var1.pred)

hist(idw_test1$var1.pred) 

st_extract(idw_test1["var1.pred"], centroid26986)

```

##create a function that runs IDW for one group and extracts the value at the population-weighted centroid
```{r 17-apply-loop-weekly}

#create a for loop that interpolates a raster for when the power plant is ON and for a specific week
#prepare what I need for the loop


wks <-sort(unique(no2_26986$week))
operating_statuses <- sort(unique(no2_26986$operating))

idw_weekly_results <- list()

idw_rasters <- list()  

row_id <- 1

for(week_id in wks){
  for(status in operating_statuses){
    
    #this filters data for unique combos of week ID and operation status
    no2group <- no2_26986 |> dplyr::filter(week == week_id, 
                                           operating == status)
  
    #skip if there aren't any points since the power plant doesn't run every week
    if (nrow(no2group) < 3 || all(is.na(no2group$no2mean))) {
      next
    }
  
   #run the IDW formula, setting idp = 2
    idw_group <- gstat::idw(formula = no2mean~1,
                         locations = no2group,
                         newdata = grid250,
                         idp = 2
                         )
    
    idw_df <- as.data.frame(idw_group["var1.pred"], xy = TRUE)

    
    ggraster <- ggplot() + geom_raster(data = idw_df,
                                       aes(x = x, y = y, fill = var1.pred)) +
      geom_sf(data = pittsfield26986, fill = NA, color = "black", linewidth = 1) +
      geom_sf(data = no2group, aes(color= "Sensors", shape = "Sensors"), size = 2, stroke = 1) +
      geom_sf(data = centroid26986, aes(color = "Population-weighted centroid", shape = "Population-weighted centroid"), size = 5, stroke = 2) +
      geom_sf(data = airport26986, aes(color = "Airport", shape = "Airport"), size = 4, stroke = 2) +
      geom_sf(data = peaker26986, aes(color = "Peaker plant", shape = "Peaker plant"), size = 4, stroke = 2) + 
      scale_fill_viridis_c(option = "turbo", na.value = NA, limits = c(0, 60)) +
      scale_color_manual(
           name = "Locations",
           values = c("Sensors" = "black","Population-weighted centroid" = "blue", "Airport" = "gray","Peaker plant" = "black")) +
      scale_shape_manual(
       name = "Locations",
        values = c(
         "Sensors" = 21,
          "Population-weighted centroid" = 4,
          "Airport" = 24,
          "Peaker plant" = 24)) +
      coord_sf() +
      labs(title = paste0("Average hourly NO2 levels for Week ", week_id, ": ", status),
           fill = "NO2 (ppb)") +
      theme_minimal()
    
    print(ggraster)
    
    # Clean file name (remove spaces)
    status_clean <- gsub(" ", "_", status)

    # Construct output file name
    png_name <- paste0("plots/ggraster_week_", week_id, "_", status_clean, ".png")

    ggsave(
        filename = png_name,
        plot     = ggraster,
        width    = 10,      # inches
        height   = 8,
        dpi      = 300      # high resolution
  )

    raster_name <- paste0("Week_", week_id, "_", status)
    idw_rasters[[raster_name]] <- idw_group 
    
    
  #extract the value of the raster at the centroid point
  centroid_no2mean <- st_extract(idw_group["var1.pred"], centroid26986)[[1]] 
  
  #assign the covariates to the extracted value for each week and operating status
  centroidno2 <- no2group |> 
    sf::st_drop_geometry() |>
    dplyr::summarise(
      week = first(week),
      operating = first(operating),
      fireworks2 = first(fireworks2),
      smoke2 = first(smoke2),
      temp = first(temp),
      humidity = first(humidity),
      windspeed = first(windspeed),
      year = first(year),
      .groups = "drop"
    )
  
  #create a new column NO2 mean based on the extracted value of the centroid 
  centroidno2$no2mean <- centroid_no2mean
  
  idw_weekly_results[[row_id]] <- centroidno2
  row_id <- row_id + 1
  }
}


centroidno2_weekly <- dplyr::bind_rows(idw_weekly_results)
head(centroidno2_weekly)

```

```{r 17-visualization-for-model2}

ggno2weeklybox <- ggplot(data = centroidno2_weekly, aes(x = factor(operating), y = no2mean)) + 
  geom_boxplot(na.rm = TRUE) +
  labs(title = "Mean hourly NO2 levels per week \nat population-weighted centroid",
       x = "Peaker plant operation status",
       y = "Concentration (ppb)") +
  theme_minimal() + 
  theme(plot.title = element_text(size = 20),
        axis.title.x = element_text(size = 18), 
        axis.title.y = element_text(size = 18),
        axis.text = element_text(size = 16))

ggsave("plots/ggno2weeklyboxplot.png", ggno2weeklybox, height = 6, width = 8, unit = "in")



```

```{r 18-run-model-2}
write.csv(centroidno2_weekly, "data/output/centroidno2_weekly.csv")

#model to include weather data. excluded windspeed because 
model2_lm1 <- lm(formula = no2mean ~ operating + week + smoke2 + 
                   fireworks2 + humidity + temp, data = centroidno2_weekly)

summary(model2_lm1)

#without weather data.
model2_lm2 <- lm(formula = no2mean ~operating + week + smoke2 + 
                   fireworks2, data = centroidno2_weekly)
summary(model2_lm2)

#check model residuals...
plot(resid(model2_lm2))
plot(model2_lm2)

plot(model2_lm1)
```

```{r 18-preparing-models-for-publication, echo = FALSE}
#tried outputting models into nice formats using stargazer
#but the outputs looked too clunky so we didn't end up using them

#library(stargazer)


#stargazer(model2_lm2, model2_lm1, type = "html",
 #         title = "Model 2 Regression Results",
  #        out = "model2_outputs.html",
   #       font.size = "normalsize",      
    #      column.sep.width = "15pt" ,
     #     depvar.labels = "Average hourly NO2 (micrograms/m3) per week")

library(webshot2)

webshot("model2_outputs.html", 
        file = "plots/model2_table.png",
        vwidth = 1000,     # width of output image
        vheight = 1000,     # height
        zoom = 3)   

```

```{r 19-get-centroid-as-shapefile}
#st_write(centroid26986, "data/output/centroid26986.shp")

#sometimes you get an error that centroid26986.shp already exists, which it does
#so I commented that code out
```



```{r 20-get-extra-numbers-for-report}
#statistics for weekly interpolated no2 levels at the centroid 
mean(centroidno2_weekly$no2mean, na.rm = TRUE)
median(centroidno2_weekly$no2mean, na.rm = TRUE)
quantile(centroidno2_weekly$no2mean, na.rm = TRUE)
sd(centroidno2_weekly$no2mean, na.rm = TRUE)

centroidno2on <- centroidno2_weekly |> filter(operating == "Operating")
mean(centroidno2on$no2mean, na.rm = TRUE)
median(centroidno2on$no2mean, na.rm = TRUE)
quantile(centroidno2on$no2mean, na.rm = TRUE)
sd(centroidno2on$no2mean, na.rm = TRUE)

centroidno2off <- centroidno2_weekly |> filter(operating == "Not operating")
mean(centroidno2off$no2mean, na.rm = TRUE)
median(centroidno2off$no2mean, na.rm = TRUE)
quantile(centroidno2off$no2mean, na.rm = TRUE)
sd(centroidno2off$no2mean, na.rm = TRUE)

```

```{r 21-calculate-distance-allendale-peake0r}
#check distance from allendale to the peaker plant
allendale26986 <- sensors26986 |> dplyr::filter(name == "Allendale Elementary School")
st_distance(allendale26986, airport26986)

#check distance from allendale to centroid 
st_distance(centroid26986, airport26986)

#st_write(allendale26986, "data/output/allendale26986.shp")

#sometimes you get an error that allendale26986.shp already exists, which it does
#so I commented the code out 
```










